Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-05-14 revision-final/paper.tex
fixed typo in discussion
fixed typo in generalization, and another near Gauguin ref

In general, multivariate glyph displays lack a natural ordering because the observations have no ordering. Similarly, there are multiple ways that variables can be mapped to glyph properties, which can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data compared to another mapping. This may have prevented the widespread adoption of glyph displays, despite some proposed ordering solutions \citep{kleiner:1981,hurley:2010}. Using glyphs with space-time data eliminates the ordering issue: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected. Additionally, climate data usually has high correlations between nearby locations and times, imposing a degree of smoothness that gives the plot an appearance of a textured  landscape, which makes it easier to digest patterns from one glyph to another. Others \citep{pickett:1988} have used glyphs on maps to display multivariate spatial data. This area of work has morphed into the field of metaphorical data displays, which create abstract landscapes of spatial data, a digression from glyph-maps. \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin. Glyph-maps for spatiotemporal data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}. Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1  and a rudimentary time series glyph display can be seen in Figure 9.12 of \citet{IPCC}.

@@

Glyphs, of course, can be more general than lines or stars. For multivariate data we may want to represent several variables ($z_1, \dots, z_p; p=$ number of variables) in the glyphs. Figure~\ref{fig:cloud} shows scatterplots of the EXPO data as icons: temperature values ($z_1$ instead of $t$) are displayed horizontally and high-cloud values ($z_2$) are displayed vertically, both locally scaled. The bivariate relationship between the two variables is explored spatially: the equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association. The right side plot displays a loess curve fit to the data instead of the individual points, sharpening the association signal realtive to the variation. Gauguin \citep{gribov:2006}, has an icon based on a histogram, giving small univariate distribution displays for each location. \citet{carr:1992} use ray glyphs to show bivariate relationships.

@@

There are two natural next directions for this work: perceptual experiments, and development of interactive graphics. Several careful studies comparing the perceptual elements of these displays, using different scale and glyph options, and in comparison with facetted heatmaps, would be recommended. For example, it appears that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs. Is this always the case? Are their certain types of periodicity that are particularly easy to spot? Rigorous perceptual studies will help guide the use of glyph-maps.

>>>>

In general, multivariate glyph displays lack a natural ordering because the observations have no ordering. Similarly, there are multiple ways that variables can be mapped to glyph properties, which can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data compared to another mapping. This may have prevented the widespread adoption of glyph displays, despite some proposed ordering solutions \citep{kleiner:1981,hurley:2010}. Using glyphs with space-time data eliminates the ordering issue: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected. Additionally, climate data usually has high correlations between nearby locations and times, imposing a degree of smoothness that gives the plot an appearance of a textured  landscape, which makes it easier to digest patterns from one glyph to another. Others \citep{pickett:1988} have used glyphs on maps to display multivariate spatial data. This area of work has morphed into the field of metaphorical data displays, which create abstract landscapes of spatial data, a digression from glyph-maps. \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software, Gauguin. Glyph-maps for spatiotemporal data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}. Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1  and a rudimentary time series glyph display can be seen in Figure 9.12 of \citet{IPCC}.

@@

Glyphs, of course, can be more general than lines or stars. For multivariate data we may want to represent several variables ($z_1, \dots, z_p; p=$ number of variables) in the glyphs. Figure~\ref{fig:cloud} shows scatterplots of the EXPO data as icons: temperature values ($z_1$ instead of $t$) are displayed horizontally and high-cloud values ($z_2$) are displayed vertically, both locally scaled. The bivariate relationship between the two variables is explored spatially: the equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association. The right side plot displays a loess curve fit to the data instead of the individual points, sharpening the association signal relative to the variation. Gauguin \citep{gribov:2006}, has an icon based on a histogram, giving small univariate distribution displays for each location. \citet{carr:1992} use ray glyphs to show bivariate relationships.

@@

There are two natural next directions for this work: perceptual experiments, and development of interactive graphics. Several careful studies comparing the perceptual elements of these displays, using different scale and glyph options, and in comparison with facetted heatmaps, would be recommended. For example, it appears that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs. Is this always the case? Are there certain types of periodicity that are particularly easy to spot? Rigorous perceptual studies will help guide the use of glyph-maps.
